Delayed effects of climate on vital rates lead to demographic divergence in Amazonian forest fragments

### 2021-08-27 review/Response to reviews.Rmd

-   Changed shape of crossbasis functions (edf changes, now not always more complex in fragments)

-   Changes in which lags are significant

-   Individual-level random effects impossible to fit due to computational limitations.
    Using a sub-sample, checked with and without invididual-level random effects.
    Changes r2, but very minimal changes to shape of smooths, edfs, and significance of smooths.
    Dropped from flowering model.

For coverletter:

we identified \_\_\_ main suggestions

1.  more in depth explanation of model (both technical and for broad audience)
2.  something about changing declaritive statements about habitat:SPEI,Lag interaction. Added caveats about limitations of DLNM approach. Expanded on this in future directions to draw attention to the importance of including interactions for demography.
3.  R\_ suggested increasing k, important suggestion. While it didn't alter the conclusions, it did require considerable work and further evaluation of results. new appendix, review rewrite

@@

(Something I'm considering adding to the discussion):

"The main limitation for detecting lagged effects with any method is data, and the specifics of data limitations have not yet been systematically explored. @tellerLinkingDemographyDrivers2016 used simulations to show that 20--25 years of data were required to detect lagged effects using their method, but this cutoff depends on the effect size and the number of climate extremes in the dataset. Additionally, there is some evidence that appropriate spatial replication could drastically reduce the number of years of data required (Aldo Compagnoni's ESA talk)."

> In some ways it seems like the analysis is adding more precision (by using monthly climate data) than the plants perceive and that would be measured in a yearly census.

Counter-intuitively, averaging weather conditions over an entire year may obscure ways plants perceive environmental variation.
For example, a drought in the wet season may have a different effect than a drought in the dry season.
This differential response to season- or month-level climate data is observable in yearly plant census.
Collapsing finer temporal scale environmental drivers to yearly averages (or a few means of arbitrarily chosen climate windows) is exactly what we avoid with this method.

@@

The difference in flowering probability between cointinuous forests and fragments is in fact significant.
We have added Table 1 that shows the model intercepts ± 84% confidence intervals, which is comparable to testing a main effect of habitat if we were able to include both habitats in the same model.

##  {#R2-7}

> Delayed effects: Since the authors fit separate models for plants in fragments and in continuous forest to allow for the crossbasis function to differ between habitats we don't know if any of the small differences reported in Fig 4c, 5c and 6c are statistically significant.
> Hence, I would recommend interpreting the results concerning the difference between fragments and CF with caution.
> I would like to see at least a justification for why representing the difference figures are meaningful.

Because we were not able to test for a significant interaction between habitat and lagged SPEI, we did not intend to suggest that our results implied statistically significant differences in the shape of the crossbasis functions between fragments and continuous forest.
We have made changes throughout the manuscript that clarify this and remove some of the stronger declarative language about differences in the effects of SPEI between continuous forest and forest fragments.
<!--# still need to do this -->

@@

> Line 301-301: This is somewhat misleading because as you describe later SPEI at time lags 15-20 and 32-36 are significant also.
> I wonder if it might be better to describe the entire Fig 4b and then the entire Fig 4a.
> What springs to my mind is that the area that is significant is much larger in in fragments compared to CF.
> It seems that in CF only a few months are significant.
> Surprisingly they fall not consistently in the dry or wet season.

##  {#R2-9}

> Line 323-325: I don't know why the growth of plants in fragments would affect the size of plant in CF.

This sentence meant to describe the difference between CF and fragments, not the absolute size in fragments.
It has been re-worded to clarify.

@@

Figure 2 shows the survivorship curve of plants tagged in 1998.
These are exact counts and therefore do not have confidence intervals.
They do not require statistics to interpret if they are different.
We think the confusion may have come from the axis scale and label ("P(survived)") which we intended to be interpreted as "proportion of plants surviving" but may have been interpreted as "probability of survival". We have changed the scale to a percentage and the axis label to "% surviving".

##  {#R2-12}

> Lines 383 ff: see comment 7

<!--# although we didn't test for a significant interaction, they have different areas of significance and different ranges of fitted values. Not sure what else to say about this--> See also our response to \@ref(R2-7).

@@

see also our response to \@ref(R2-7).

##  {#R2-23}

> Line 499: This is confusing.
> If you measure size every year than you calculate growth as the change in size from one year to the next, don't you?

That's correct.
This sentence meant to point out that size is usually measured in yearly censuses while weather data can be more easily collected at monthly or daily timescales.
We've edited this sentence to read: "The ability to identify size-specific lagged responses may be especially complicated given size is rarely measured at the same time scale (e.g. monthly) as climate drivers."

### 2021-08-27 doc/paper.Rmd
add reference to fig 2

*Survival:* Across all plots, the proportion of plants surviving was lowest in the 2003--2004 transition year (93% survival in both continuous forest and forest fragments).
This coincided with droughts in both the 2003 and 2004 rainy seasons (Figure \@ref(fig:eda)b,d) and was preceded by a drop in average plant size in the 2002--2003 transition year (i.e. negative growth in Figure \@ref(fig:eda)a).
There was no significant difference in survival between continuous forest and 1 ha fragments (Table \@ref(tab:intercept-table)).
Survival in both habitats was significantly affected by size in the previous year (Table \@ref(tab:results-table)) with larger plants having much greater survival probabilities (Figure \@ref(fig:size-plot)b).
The survival probability of large plants approached 1 in both habitat types, but the smallest plants had higher survival in 1 ha fragments (although 95% confidence intervals overlap at all sizes).

@@

*Survival:* For 1-ha fragments, there was a significant effect on survival of SPEI in the preceding 0--16 months.
The highest survival was near SPEI of 0, with mortality increasing as conditions became either drier or wetter (i.e., as SPEI values became increasingly negative or positive, respectively; Figure \@ref(fig:surv-spei)b).
In contrast, only recent conditions (0--4 month delay) has a significant effect on survival in continuous forest (Figure \@ref(fig:surv-spei)a).
The short-term effects of SPEI on survival in continuous forest were also unidirectional---the probability of survival was highest in wet conditions and declined, albeit only slightly, with increasingly negative values of SPEI (i.e., as droughts became more severe; Figure \@ref(fig:surv-spei)a).
Finally, plants in both habitat types showed an increase in survival probability with very high SPEI values (i.e., extremely high precipitation) at a lag time of 32--36 months.
It should be noted, however, that only the first transition year of census data (1998--1999) met these conditions.
We compared the effects of SPEI history in continuous forest and fragments by subtracting the fitted values in Figure \@ref(fig:surv-spei)b from Figure \@ref(fig:surv-spei)a to produce Figure \@ref(fig:surv-spei)c.
This shows that in average conditions (SPEI = 0), there is little difference in survival probability between continuous forest and forest fragments (Figure \@ref(fig:surv-spei)c).
However, under extreme conditions, survival probability is higher in continuous forest by up to 0.042.

>>>>

*Survival:* Across all plots, the proportion of plants surviving was lowest in the 2003--2004 transition year (93% survival in both continuous forest and forest fragments).
This coincided with droughts in both the 2003 and 2004 rainy seasons (Figure \@ref(fig:eda)b,d) and was preceded by a drop in average plant size in the 2002--2003 transition year (i.e. negative growth in Figure \@ref(fig:eda)a).
There was no significant difference in survival between continuous forest and 1 ha fragments (Table \@ref(tab:intercept-table)).
Although differences in survival probability are small and not statistically significant, these slight differences can be seen to compound over time in the survivorship of plants labeled in the first year of the surveys (Figure \@ref(fig:surv-curve)).
Survival in both habitats was significantly affected by size in the previous year (Table \@ref(tab:results-table)) with larger plants having much greater survival probabilities (Figure \@ref(fig:size-plot)b).
The survival probability of large plants approached 1 in both habitat types, but the smallest plants had higher survival in 1 ha fragments (although 95% confidence intervals overlap at all sizes).

@@

*Survival:* For 1-ha fragments, there was a significant effect on survival of SPEI in the preceding 0--16 months.
The highest survival was near SPEI of 0, with mortality increasing as conditions became either drier or wetter (i.e., as SPEI values became increasingly negative or positive, respectively; Figure \@ref(fig:surv-spei)b).
In contrast, only recent conditions (0--4 month delay) has a significant effect on survival in continuous forest (Figure \@ref(fig:surv-spei)a).
The short-term effects of SPEI on survival in continuous forest were also unidirectional---the probability of survival was highest in wet conditions and declined, albeit only slightly, with increasingly negative values of SPEI (i.e., as droughts became more severe; Figure \@ref(fig:surv-spei)a).
Finally, plants in both habitat types showed an increase in survival probability with very high SPEI values (i.e., extremely high precipitation) at a lag time of 32--36 months.
It should be noted, however, that only the first transition year of census data (1998--1999) met these conditions.
We compared the effects of SPEI history in continuous forest and fragments by subtracting the fitted values in Figure \@ref(fig:surv-spei)b from Figure \@ref(fig:surv-spei)a to produce Figure \@ref(fig:surv-spei)c.
This shows that in average conditions (SPEI = 0), there is little difference in survival probability between continuous forest and forest fragments (Figure \@ref(fig:surv-spei)c).
However, under extreme conditions, survival probability is higher in continuous forest compared to fragments by up to 0.042.
